Typed Revision Control

### 2024-06-11 mss.tex
Varia

A revision control system\footnote{Also “version control system” (VCS); “source control system”.} is responsible for tracking the state and history of assets and asset changes within a particular continuity.  It may also manage the production of assets through a build system, as used for instance by a continuous integration/continuous deployment (CI/CD) workflow.  Trivially, an RCS is a logical filesystem:  a database for managing files.  (Conventional RCSs are agnostic to the underlying disk hardware, however.\footnote{Compare other logical file systems:  network file systems like Microsoft Server Message Block (SMB) drives or distributed file systems like the Hadoop Distributed File System (HDFS) or the InterPlanetary File System (IPFS).})  Computer revision control systems have developed from the 1960s onwards once hard drives permitted files to be stored on tape rather than simply as punchcard files.  Succeeding generations have introduced new approaches and features, converging more or less on the broad functionality today afforded by Git \citep{Torvalds2005} in its elaborated form as a distributed source control system.\footnote{Notably, Git has seemingly failed to fulfill its original promise of being a fully decentralized RCS:  most users prefer to use the affordances of centralized Git services such as GitHub or GitLab.}  However, although some modest interest has been expressed for a more complete notion of type in revision control, no current system besides Urbit's Clay fully implements a typed revision-controlled filesystem.  In this article, we explore the approach of current RCSs and elaborate Clay's contributions to the ongoing task of information management and auditable, reproducible source builds.
% Torvalds2005, https://github.com/git/git/commit/e83c5163316f89bfbde7d9ab23ca2e25604af290

\section{A History of Type in RCSs}

The problem of type in revision control has been of both academic and practical interest for decades \citep{Perry1987}.  Essentially, can file artifacts be organized and classified in such a way as to make them susceptible of comparison across their history?  The simplest arrangement which is commonly made is to designate a file as either plaintext or binary.  A plaintext file can be updated by specifying the character offset and run length to be replaced (an example of a “diff”, or difference between two examples of text; typically these are resolved at the line level rather than character level).  Plaintext files, whether ASCII or some form of UTF, are defined by a regular format with well-understood and predictable symbol widths and offsets.  For instance, a little-endian representation of the text \texttt{'clay'} would be reproduced in ASCII hexadecimal as \texttt{0x7961.6c63} and ASCII binary as \texttt{0b111.1001.0110.0001.0110.1100.0110.0011}.  (In binary, each character has a leading zero; that is, each entry is seven bits wide plus a 1-bit zero header.)  In contrast, binary files (which is a byword for “everything else”) have arbitrary layout and offset; it may be difficult to determine how and why to compare two revisions of any given file.
% Perry1987, https://users.ece.utexas.edu/~perry/work/papers/icse9b.pdf

All modern RCSs support plaintext and binary representations of files.  Plaintext diffing at the level of line changes is straightforward; the Hunt–McIlroy algorithm is often employed \citep{Hunt1976}.\footnote{An implementation in Hoon is available at \github{https://github.com/sigilante/diff}.}  However, the most popular revision control systems (such as Subversion and Git) have included only minimal direct support for binary diffs.  Some systems only support storing successive file copies of a binary file; that is, the diff is the entire deletion of the old file and addition of the new file (e.g. CVS).  Other systems may support binary file diffs by granular block and offset.  Some RCSs support diffing particular binary files using a custom automatic merge tool.  For example, the \texttt{textconv} tool converts each version of the file to a plaintext representation and compares those representations.  (To show a diff for a changes in a \texttt{pdf} file, a configured Git instance could compare the text output of \texttt{pdfinfo} in each case as a proxy.  This is generally legible as metadata but does not of itself include sufficient information to reproduce or reverse the intervening changes; the underlying RCS still stores the entire binary in its conventional representation.)  In any case, type beyond file extension is not generally supported by RCSs; they cannot meaningfully merge files classified as binary.\footnote{As always, there are exceptions to the rules.  Git, for instance, can use a “binary diff” format on exact preimages.}
%  Hunt, James W.; McIlroy, M. Douglas (June 1976). "An Algorithm for Differential File Comparison" (PDF). Computing Science Technical Report. 41. Bell Laboratories.  http://www.cs.dartmouth.edu/~doug/diff.pdf

A general-purpose notion of file type is not isomorphic to conventional type in programming language theory.  For instance, consider a JSON file.  The following are equivalent as JSON artifacts:

\begin{lstlisting}
{
  "name": "Alice",
  "age": 30
}
\end{lstlisting}

\begin{lstlisting}
{
  "age": 30,
  "name": "Alice"
}
\end{lstlisting}

\begin{lstlisting}
{ "name": "Alice", "age": 30, }
\end{lstlisting}

Indeed, since JSON is structurally agnostic to syntactic whitespace, an infinite number of equivalent files could be constructed.  Clearly as plaintext representations, each of these differ, yet they are fully equivalent as JSON artifacts.  An RCS which can recognize a JSON input can meaningfully store them as equivalent, while an RCS that makes only the trivial distinction between plaintext and binary must consider them all to be different.  Furthermore, a JSON-aware RCS could meaningfully merge these files whereas a plaintext/binary RCS could not.  Finally, a JSON-aware RCS could yield a different plaintext output when asked to produce the file, as it could normalize the whitespace to a canonical form.  This may be considered by other systems to be an ``incorrect'' round trip.

Thus we must consider two intersecting notions of file type:
\begin{enumerate}
  \item The file type as understood by the system, typically based on bitwise reproducibility (e.g. plaintext, binary).
  \item The file type as understood conceptually, typically based on structural relationships (e.g. JSON, XML/HTML, Word \texttt{docx}, etc.).
\end{enumerate}

\noindent
The employment of an intermediate representation allows us to track changes to files and manage their history—state—more effectively.  Consider the popular JavaScript framework React, aspects of which were developed to resolve difficulties with slow repainting in browsers (for a popular explanation, see \citet{Arora2017}).  React's virtual DOM system acts as an intermediate representation between the actual DOM and the developer's code.  When a developer updates the virtual DOM, the framework compares the new virtual DOM with the previous virtual DOM and updates the real DOM only if necessary.  This allows React to optimize the rendering process and improve performance.  It also means that we can consider the structural representation of the DOM object as an intermediate representation, and the production and application of the diff as a type-aware RCS operation.
%TODO don't use this one Arora2017 https://medium.com/@aankit35/what-problem-does-react-js-solve-for-software-developers-a06cf50be28a
Higher-level interfaces like Microsoft Word's “Track Changes” feature \citep{MS2024} similarly resolves diffs across text and formatting within a document, an artifact processed from a file into a user-friendly representation.  While the “Track Changes” user interface is optimized for interactivity, conceptually it is quite similar to conventional merge tools from RCSs, and de facto resolves the same sorts of diffs for its particular file type.  (Indeed, one can imagine a Word-aware CLI RCS that could meaningfully merge changes to a Word document, rather than simply storing successive copies of the file.)
%MS2024 https://support.microsoft.com/en-us/office/track-changes-in-word-197ba630-0f5f-4a8e-9a77-3712475e806a
Thus a significant part of managing typed file data is deciding what the input, output, and diffs should look like, and the remainder is systematic application of these rules to the data.

Another possible move is to abstract away from the file as hte fundamental artifact being tracked.  While Clay does not fully commit to this, in a concession to its POSIX-compliant host OS,

@@

\noindent
In this scheme, contemporary Git-style RCS is a ``weakly-typed'' RCS:  the system is aware of versions and diffs, but typically not below the file or line level.\footnote{Oddly, Perry did not call out weak typing in his enumeration, although he did address it elsewhere in the article.}  Perry was concerned with the behavior and representation of semantic objects, rather than files or data:

\begin{quote}
  Version equivalence in this type of version control mechanism is defined in terms of syntactic equivalence:  data objects are equivalent if their types or structures are equivalent; operations and modules are equivalent if their signatures are equivalent.
\end{quote}

\noindent
This scheme leads to a notion of version compatibility—diffs, if you will—that is analogous to type variance in classical type systems. TODO

https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=e572edf81da7b3130abbb9050a4854993a6124d8
https://users.ece.utexas.edu/~perry/work/papers/icse9b.pdf
https://urbit.org/blog/toward-a-new-clay
https://www.microsoft.com/en-us/research/wp-content/uploads/2015/05/beyond-open-source.pdf
https://stackoverflow.com/questions/73797/how-do-i-tell-subversion-to-treat-a-file-as-a-binary-file
https://chat.openai.com/c/b031b6f7-d6d8-4503-98bc-370cafbfe31d

From page 367 of the [Subversion book](http://svnbook.red-bean.com/en/1.5/svn-book.pdf)

> In the most general sense, Subversion handles binary files more gracefully than CVS does. Because CVS uses RCS, it can only store successive full copies of a changing binary file. Subversion, however, expresses differences between files using a binary differencing algorithm, regardless of whether they contain textual or binary data. That means all files are stored differentially (compressed) in the repository.
> 
> CVS users have to mark binary files with -kb flags to prevent data from being garbled (due to keyword expansion and line-ending translations). They sometimes forget to do this.
> 
> Subversion takes the more paranoid route. First, it never performs any kind of keyword or line-ending translation unless you explicitly ask it to do so (see the section called “Keyword Substitution” and the section called “End-of-Line Character Sequences” for more details). By default, Subversion treats all file data as literal byte strings, and files are always stored in the repository in an untranslated state.
> 
> Second, Subversion maintains an internal notion of whether a file is “text” or “binary” data, but this notion is only extant in the working copy. During an svn update, Subversion will perform contextual merges on locally modified text files, but will not attempt to do so for binary files.
> 
> To determine whether a contextual merge is possible, Subversion examines the svn:mime-type property. If the file has no svn:mime-type property, or has a MIME type that is textual (e.g., text/*), Subversion assumes it is text. Otherwise, Subversion assumes the file is binary. Subversion also helps users by running a binary-detection algorithm in the svn import and svn add commands. These commands will make a good guess and then (possibly) set a binary svn:mime-type property on the file being added. (If Subversion guesses wrong, the user can always remove or hand-edit the property.)

@@

Urbit provides the Clay vane at \texttt{/sys/vane/clay} to manage Unix-style files and source builds.  Clay acts as both the Unix-like filesystem and the source control vane.  While Hoon itself (\texttt{/sys/hoon}) is responsible to compile a source noun into a Hoon AST and ultimately into executable Nock, the actual construction of source input from files is delegated to the \texttt{++ford} arm of Clay.  This build process includes compiling the correct import files (such as libraries and shared structure files).  Clay is also tightly coupled to the userspace vane Gall since system updates and agent updates both occur via Clay.\footnote{We omit discussion of several other notable properties of Clay as a single-level store:  referential transparency, global addressability, and event-level persistence, for instance.}

To motivate how Clay functions as a revision control filesystem, we need to briefly examine some major data structures utilized by Clay.  At the highest level, files are organized into “desks”, described in previous literature as being analogous to Git branches but really more like Git repositories.  A desk is a self-contained continuity of files; its state is the result of a history of commits.  Desks may not include links to off-desk resources.  Each desk must expose a few files at definite paths; primarily \texttt{/mar} for the marks used to read the desk contents.\footnote{While Arvo, the vanes, and userspace tools look at other definite locations, in principle one could construct a desk of arbitrary URL-safe paths as long as they are legible to Clay's build process via \texttt{/mar}.}

Ultimately, Arvo and Clay deal in nouns.  A noun is a binary tree of unsigned integers.  These may be evaluated as Nock formulas or manipulated as data.  A noun is in a sense a noumenon; it is formally agnostic to the underlying hardware and is almost always accessed via some representation path (i.e. as phenomenon).  At least, to leave off of philosophy, a noun is an intermediate representation which is common to all Nock-legible data.

Urbit has an immutable data model; the system state of Arvo is the unique result of the events in its history, stored as its event log.  If a noun changes, it is the result of 
(This yields very nice properties of referential transparency and has some ramifications in library management; see \citet{Blackman2024} for more details.)

commits
diffs
etc.

@@

A mark may be considered an executable media type.\footnote{A media type (formerly MIME type) is a tag identifier for the intended data format of a particular data entry.}  A mark essentially describes how to map a data representation to a noun, and from a noun back out to a data representation.  This conversion may be trivial (e.g. the binary storage of an audio file) or sophisticated (from a text stream to a linked-list UTF representation).  Data are accessed via a particular mark, and the appropriate conversion routine is invoked.\footnote{This is commonly used with JSON and the Eyre HTTP server vane, for instance.}  Clay is capable of searching for transition paths between mark representations, and in principle permits conversion between all compatible representations (if there were a path from Markdown \texttt{md} to rich text \texttt{rtf} to plaintext \texttt{txt} in marks, then the conversion is automatically supported by Clay).

Custom filetype

\subsection[Shortcomings]{Shortcomings of Clay}

@@

\begin{quote}
  A mark is a term (symbol) which is the Urbit equivalent of a MIME type, if MIME types were names of typed validation functions.  \ldots{} Like a MIME type, the mark is just a label. There is no way to guarantee that the sender and receiver agree on what this label means.  A noun which doesn't normalize to itself is a packet drop.  \citep[pp. 51–52]{Whitepaper}
\end{quote}

@@

\begin{enumerate}
  \item  Implementing RCS behavior at the file system layer seems to be incorrect for Urbit (that is, too much of a sop to Unix).  trying to do revision control at base layer, but file system is wrong layer (you just want history of actions/tx log rather than full state); see ~migrev codebases for clay state
  \item  Desks seem to be the wrong abstraction in practice.  The theory of the desk doesn't account for actual distribution patterns people would like to use; we want one desk but one spot in file tree is capable of acting as a desk (chroot analysis); overuse of desks makes Urbit not feel as much like a filesystem you can explore
  \item  The application system (userspace) and the build system and filesystem should be co-located.  Currently, Clay and Gall are deeply entangled but must interact via a somewhat awkward message-passing interface.  This leads to a number of ergonomic issues, such as the need to suspend a desk in order to suspend an agent.  Employing paths rather than simple tags would lead to a desire to constrain what can exist at certain paths, which of course relates to the definition of files as data structures.
%3. application system and files should be co-located, message passing b/w C/G is weird; you have to suspend a desk, even if you want to suspend an agent, etc. C/G trying to marry each other but operate through back and forth; (Were there ever any docs written about the old Hume concept?); paths instead of tags leads you to want to constrain what can exist at certain paths; rest of shrubbery is solving for that problem
\end{enumerate}

The future of Clay likely sees it being restricted to source and build management, rather than expansion to a more fully-featured filesystem.  (In this contingency, the userspace management vane likely takes over conventional file storage, Gall or a successor.)  However, in its current instantiation, Clay exhibits some notable characteristics.

@@

\printbibliography
\end{document}

>>>>

A revision control system\footnote{Also “version control system” ({\sc vcs}); “source control system”.} is responsible for tracking the state and history of assets and asset changes within a particular continuity.  It may also manage the production of assets through a build system, as used for instance by a continuous integration/continuous deployment (CI/CD) workflow.  Trivially, an {\sc rcs} is a logical filesystem:  a database for managing files.  (Conventional {\sc rcs}s are agnostic to the underlying disk hardware, however.\footnote{Compare other logical file systems:  network file systems like Microsoft Server Message Block ({\sc smb}) drives or distributed file systems like the Hadoop Distributed File System ({\sc hdfs}) or the InterPlanetary File System ({\sc ipfs}).})  Computer revision control systems have developed from the 1960s onwards once hard drives permitted files to be stored on tape rather than simply as punchcard files.  Succeeding generations have introduced new approaches and features, converging more or less on the broad functionality today afforded by Git \citep{Torvalds2005} in its elaborated form as a distributed source control system.\footnote{Notably, Git has seemingly failed to fulfill its original promise of being a fully decentralized {\sc rcs}:  most users prefer to use the affordances of centralized Git services such as GitHub or GitLab.}  However, although some modest interest has been expressed for a more complete notion of type in revision control, no current system besides Urbit's Clay fully implements a typed revision-controlled filesystem.  In this article, we explore the approach of current {\sc rcs}s and elaborate Clay's contributions to the ongoing task of information management and auditable, reproducible source builds.
% Torvalds2005, https://github.com/git/git/commit/e83c5163316f89bfbde7d9ab23ca2e25604af290

\section{A History of Type in {\sc rcs}s}

The problem of type in revision control has been of both academic and practical interest for decades \citep{Perry1987}.  Essentially, can file artifacts be organized and classified in such a way as to make them susceptible of comparison across their history?  The simplest arrangement which is commonly made is to designate a file as either plaintext or binary.  A plaintext file can be updated by specifying the character offset and run length to be replaced (an example of a “diff”, or difference between two examples of text; typically these are resolved at the line level rather than character level).  Plaintext files, whether {\sc ascii} or some form of {\sc utf}, are defined by a regular format with well-understood and predictable symbol widths and offsets.  For instance, a little-endian representation of the text \texttt{'clay'} would be reproduced in {\sc ascii} hexadecimal as \texttt{0x7961.6c63} and {\sc ascii} binary as \texttt{0b111.1001.0110.0001.0110.1100.0110.0011}.  (In binary, each character has a leading zero; that is, each entry is seven bits wide plus a 1-bit zero header.)  In contrast, binary files (which is a byword for “everything else”) have arbitrary layout and offset; it may be difficult to determine how and why to compare two revisions of any given file.
% Perry1987, https://users.ece.utexas.edu/~perry/work/papers/icse9b.pdf
Many {\rsc rcs}s simply treat all registered data as byte streams.

All modern {\sc rcs}s support plaintext and binary representations of files.  Plaintext diffing at the level of line changes is straightforward; the Hunt–McIlroy algorithm is often employed \citep{Hunt1976}.\footnote{An implementation in Hoon is available at \github{sigilante/diff}.}  However, the most popular revision control systems (such as Subversion and Git) have included only minimal direct support for binary diffs.  Some systems only support storing successive file copies of a binary file; that is, the diff is the entire deletion of the old file and addition of the new file (e.g. {\sc cvs}).  Other systems may support binary file diffs by granular block and offset.  Some {\sc rcs}s support diffing particular binary files using a custom automatic merge tool.  For example, the \texttt{textconv} tool converts each version of the file to a plaintext representation and compares those representations.  (To show a diff for a changes in a {\sc pdf} file, a configured Git instance could compare the text output of \texttt{pdfinfo} in each case as a proxy.  This is generally legible as metadata but does not of itself include sufficient information to reproduce or reverse the intervening changes; the underlying {\sc rcs} still stores the entire binary in its conventional representation.)  In any case, type beyond file extension is not generally supported by {\sc rcs}s; they cannot meaningfully merge files classified as binary.\footnote{As always, there are exceptions to the rules.  The Git {\sc rcs}, for instance, can use a “binary diff” format on exact preimages.  The Subversion {\sc rcs} does provide an elaborated {\sc mime} support system, wherein users can assign each file a {\sc mime} type that Subversion can use in applying and displaying diffs.  However, in practice this only results in the classic text/binary dichotomy with slightly more granularity.  ``Subversion treats all file data as literal byte strings, and files are always stored in the repository in an untranslated state'' \citep{Collins2016}.
% Collins2016 https://svnbook.red-bean.com/en/1.8/svn.forcvs.binary-and-trans.html
}

%  Hunt, James W.; McIlroy, M. Douglas (June 1976). "An Algorithm for Differential File Comparison" (PDF). Computing Science Technical Report. 41. Bell Laboratories.  http://www.cs.dartmouth.edu/~doug/diff.pdf

A general-purpose notion of file type is not isomorphic to conventional type in programming language theory.  For instance, consider a {\sc json} file.  The following are equivalent as {\sc json} artifacts:

\begin{lstlisting}[style=listingcode]
{
  "name": "Alice",
  "age": 30
}
\end{lstlisting}

\begin{lstlisting}[style=listingcode]
{
  "age": 30,
  "name": "Alice"
}
\end{lstlisting}

\begin{lstlisting}[style=listingcode]
{ "name": "Alice", "age": 30, }
\end{lstlisting}

Indeed, since {\sc json} is structurally agnostic to syntactic whitespace, an infinite number of equivalent files could be constructed.  Clearly as plaintext representations, each of these differ, yet they are fully equivalent as {\sc json} artifacts.  An {\sc rcs} which can recognize a {\sc json} input can meaningfully store them as equivalent, while an {\sc rcs} that makes only the trivial distinction between plaintext and binary must consider them all to be different.  Furthermore, a {\sc json}-aware {\sc rcs} could meaningfully merge these files whereas a plaintext/binary {\sc rcs} could not.  Finally, a {\sc json}-aware {\sc rcs} could yield a different plaintext output when asked to produce the file, as it could normalize the whitespace to a canonical form.  This may be considered by other systems to be an ``incorrect'' round trip.

Thus we must consider two intersecting notions of file type:
\begin{enumerate}
  \item The file type as understood by the system, typically based on bitwise reproducibility (e.g. plaintext, binary).
  \item The file type as understood conceptually, typically based on structural relationships (e.g. {\sc json}, {\sc xml}/{\sc html}, Word \texttt{docx}, etc.).
\end{enumerate}

\noindent
The employment of an intermediate representation allows us to track changes to files and manage their history—state—more effectively.  Consider the popular JavaScript framework React \citep{React}, aspects of which were developed to resolve difficulties with slow repainting in browsers (for a popular explanation, see \citet{Arora2017}).  React's virtual Document Object Model ({\sc dom}) system acts as an intermediate representation between the actual {\sc dom} and the developer's code.  When a developer updates the virtual {\sc dom}, the framework compares the new virtual {\sc dom} with the previous virtual {\sc dom} and updates the real {\sc dom} only if necessary.  This allows React to optimize the rendering process and improve performance.  It also means that we can consider the structural representation of the {\sc dom} object as an intermediate representation, and the production and application of the diff as a type-aware {\sc rcs} operation.
%TODO don't use this one Arora2017 https://medium.com/@aankit35/what-problem-does-react-js-solve-for-software-developers-a06cf50be28a
%React https://react.dev/
Higher-level interfaces like Microsoft Word's “Track Changes” feature \citep{MS2024} similarly resolves diffs across text and formatting within a document, an artifact processed from a file into a user-friendly representation.  While the “Track Changes” user interface is optimized for interactivity, conceptually it is quite similar to conventional merge tools from {\sc rcs}s, and de facto resolves the same sorts of diffs for its particular file type.  (Indeed, one can imagine a Word-aware command-line {\sc rcs} that could meaningfully merge changes to a Word document, rather than simply storing successive copies of the file.)
%MS2024 https://support.microsoft.com/en-us/office/track-changes-in-word-197ba630-0f5f-4a8e-9a77-3712475e806a
Thus a significant part of managing typed file data is deciding what the input, output, and diffs should look like, and the remainder is systematic application of these rules to the data.

@@

\noindent
In this scheme, contemporary Git-style {\sc rcs} is a ``weakly-typed'' {\sc rcs}:  the system is aware of versions and diffs, but typically not below the file or line level.\footnote{Oddly, Perry did not call out weak typing in his enumeration, although he did address it elsewhere in the article.}  Perry was concerned with the behavior and representation of semantic objects, rather than files or data:

\begin{quote}
  Version equivalence in this type of version control mechanism is defined in terms of syntactic equivalence:  data objects are equivalent if their types or structures are equivalent; operations and modules are equivalent if their signatures are equivalent.  \citep{Perry1987}
\end{quote}

\noindent
Strongly-typed version control requires coupling to the build system, since the system must be aware of the structure of the data in order to apply diffs.  (Indeed, \citet{Blackman2024} discusses one solution to the linking problem which is well-suited to Urbit's global namespace.)

@@

\subsection{Architecture and Revision Control}

Urbit provides the Clay vane at \lstinline[style=inlinecode]{/sys/vane/clay} to manage Unix-style files and source builds.  Clay acts as both the Unix-like filesystem and the source control vane.  While Hoon itself (\lstinline[style=inlinecode]{/sys/hoon}) is responsible to compile a source noun into a Hoon abstract syntax tree ({\sc ast}) and ultimately into executable Nock, the actual construction of source input from files is delegated to the \lstinline[style=inlinecode]{++ford} arm of Clay.  This build process includes compiling the correct import files (such as libraries and shared structure files).  Clay is also tightly coupled to the userspace vane Gall since system updates and agent updates both occur via Clay.\footnote{We omit discussion of several other notable properties of Clay as a single-level store:  referential transparency, global addressability, and event-level persistence, for instance.}

To motivate how Clay functions as a revision control filesystem, we need to briefly examine some major data structures utilized by Clay.  At the highest level, files are organized into “desks”, described in previous literature as being analogous to Git branches but really more like Git repositories.  A desk is a self-contained continuity of files; its state is the result of a history of commits.  Desks may not include links to off-desk resources.  Each desk must expose a few files at definite paths; primarily \lstinline[style=inlinecode]{/mar} for the marks used to read the desk contents.\footnote{While Arvo, the vanes, and userspace tools look at other definite locations, in principle one could construct a desk of arbitrary {\sc url}-safe paths as long as they are legible to Clay's build process via \lstinline[style=inlinecode]{/mar}.}

Ultimately, Arvo and Clay deal in nouns.  A noun is a binary tree of unsigned integers.  These may be evaluated as Nock formulas or manipulated as data.  A noun is formally agnostic to the underlying hardware and is almost always accessed via some representation path (i.e. as phenomenon).  A noun serves as a universally legible intermediate representation for all data and code.

Urbit has an immutable data model; the system state of Arvo is the unique result of the events in its history, stored as its event log.  If a noun changes, it is the result of definite discrete changes to the system state.  This yields very nice properties of referential transparency and has some ramifications in library management; see \citet{Blackman2024} for more details.

Typically, a noun is altered either directly via subject-modi\-fying code (\lstinline[style=inlinecode]{%~} censig, \lstinline[style=inlinecode]{=.} tisdot) or via a mark transformation, on which more later.

@@

A mark may be considered an executable {\sc mime} type.\footnote{A media type (formerly {\sc mime} type) is a tag identifier for the intended data format of a particular data entry. We will use “{\sc mime} type” as this terminology remains in common use.}  A mark essentially describes how to map a data representation to a noun, and from a noun back out to a data representation.  This conversion may be trivial (e.g. the binary storage of an audio file) or sophisticated (from a text stream to a linked-list {\sc utf} representation).  Data are accessed via a particular mark, and the appropriate conversion routine is invoked.\footnote{This is commonly used with {\sc json} and the Eyre {\sc http} server vane, for instance.}  Clay is capable of searching for transition paths between mark representations, and in principle permits conversion between all compatible representations (if there were a path from Markdown \texttt{md} to rich text \texttt{rtf} to plaintext \texttt{txt} in marks, then the conversion is automatically supported by Clay).

\begin{quote}
  A mark is a term (symbol) which is the Urbit equivalent of a {\sc mime} type, if {\sc mime} types were names of typed validation functions.  \citep[p. 51]{Whitepaper}
\end{quote}

Unix filesystem mounting and committing involves synchronizing a desk's state with a Unix logical representation of the files involved.  The actual file data are imported as \lstinline[style=inlinecode]{%mime}-typed nouns and then converted to their target mark as file type.

Each mark defines several standard arms:

\begin{enumerate}
  \item  \lstinline[style=inlinecode]{++grab} converts from other marks (by arm name) to a given mark.
  \item  \lstinline[style=inlinecode]{++grow} converts from the mark to other marks.
  \item  \lstinline[style=inlinecode]{++grad} defines the diffs applicators for the mark.
\end{enumerate}

\noindent
Marks do not need to be symmetrical or round-trip.  A {\sc json} input may yield an equivalent {\sc json} output but altered by structural whitespace.  Clay is also proactive in searching for possible mark paths—if an immediate conversion is not available, then a mutually available intermediary form may be used.  To implement a custom filetype, all that is required is the production of a suitable mark file.

While Clay has at times supported diffs, it currently simply stores the entire file at each commit.  (No technical limitation is in place; this was merely a stylistic change while Urbit data storage is relatively small and it is expected to be reverted as larger loom sizes are supported.)

A desk is a filesystem continuity, and represents the main way that data are synchronized between Urbit instances (ships).  The state of a desk is a history of sequential commits, with each commit after the first having at least one parent commit.  (Desks are often analogized to Git repositories.)

TODO

diffs
etc.
yokes domes

Structurally, Clay

support for merge strategies etc.

story

As a vane of Arvo, Clay lives ``on Mars''; that is, it describes a logical filesystem inside of the Urbit ship but does not directly map its nouns to physical hardware.  The particular implementation used to store the total state of Arvo, which contains Clay's state including file data, depends on the runtime.  At the time of writing, the Vere runtime uses the Lightning Memory-Mapped Database Manager ({\sc lmdb}) which is a B-tree representation for a key-value store \citep{LMDB}.
%LMDB http://www.lmdb.tech/doc/

In principle, as we have intimated, an {\rcs} may abstract away from the file as the fundamental artifact being tracked.  While Clay does not fully commit to this, in a concession to its {\sc posix}-compliant host OS, it gestures towards the possibility of conceiving of build sources and artifacts as objects (cores) other than files.

Clay is also directly responsible for building code.  Formerly, Urbit built code from source files using the Ford vane.  In 2020, the core developers realized that integrating the build system directly into the {\sc rcs} solved a number of problems with updating the system correctly.  Updates to the system should be atomic (complete or fail in one transaction); self-contained (no implicit dependencies or dependencies on previous system states); and ordered (sequenced within the system stack).  By coupling the build system into the {\sc rcs}, the system can be updated in a single transaction, with the build system ensuring that the system is always in a consistent state.  In a typed {\sc rcs}, the compiler or build system must be always at hand to process diffs in artifacts.  Although of relatively minor consequence to userspace application developers, the integration of \lstinline[style=inlinecode]{++ford} into Clay transforms the system closer into an integrated {\sc rcs} for generic nouns.

\subsection{Shortcomings}

@@

\begin{quote}
  Like a {\sc mime} type, the mark is just a label. There is no way to guarantee that the sender and receiver agree on what this label means.  A noun which doesn't normalize to itself is a packet drop.  \citep[pp. 51–52]{Whitepaper}
\end{quote}

@@

\begin{enumerate}
  \item  Implementing {\sc rcs} behavior at the file system layer seems to be incorrect for Urbit (that is, too much of a sop to Unix).  trying to do revision control at base layer, but file system is wrong layer (you just want history of actions/tx log rather than full state); see ~migrev codebases for clay state
  \item  Desks seem to be the wrong abstraction in practice.  The theory of the desk doesn't account for actual distribution patterns people would like to use; we want one desk but one spot in file tree is capable of acting as a desk (chroot analysis); overuse of desks makes Urbit not feel as much like a filesystem you can explore
  \item  The application system (userspace) and the build system and filesystem should be co-located.  Currently, Clay and Gall are deeply entangled but must interact via a somewhat awkward message-passing interface.  This leads to a number of ergonomic issues, such as the need to suspend a desk in order to suspend an agent.  Employing paths rather than simple tags would lead to a desire to constrain what can exist at certain paths, which of course relates to the definition of files as data structures.
\end{enumerate}

One possible future for Clay sees it being restricted to source and build management, rather than expansion to a more fully-featured filesystem.  In this contingency, the userspace management vane may take over conventional file storage, as Gall or a successor.

Alternatively, the build system, {\sc rcs}, and application engine fuse into a single userspace vane, which is a possible endpoint of the ``shrubbery'' project.  Mark cores would be replaced with simpler conversion rules permitting only straightforward type casts between nouns as \lstinline[style=inlinecode]{$-(from to)} gates.

\section{Conclusion}

In its current instantiation, Clay exhibits some notable characteristics as a typed revision control system.  It is capable of tracking file changes at a structural level, permitting meaningful diffs and merges for files of the same mark.  It is also capable of converting between marks, permitting a more flexible approach to file type than the conventional text/binary dichotomy.  However, the future of userspace and code building is still in flux and may see a reworking of the Clay vane to better suit the needs of developers and users.

@@

%https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=e572edf81da7b3130abbb9050a4854993a6124d8
%https://users.ece.utexas.edu/~perry/work/papers/icse9b.pdf
%https://urbit.org/blog/toward-a-new-clay
%https://www.microsoft.com/en-us/research/wp-content/uploads/2015/05/beyond-open-source.pdf
%https://stackoverflow.com/questions/73797/how-do-i-tell-subversion-to-treat-a-file-as-a-binary-file
%https://chat.openai.com/c/b031b6f7-d6d8-4503-98bc-370cafbfe31d

\selectlanguage{USenglish}
\printbibliography
\end{document}
